Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-05-13 Paper/manuscript_v2/manuscript_07_02.tex
More minor edits

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions. However they entail a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis function approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance and practical implementation of the method in relation to key factors such as the number of basis functions, domain of the prediction space, and smoothness of the latent function. We provide intuitive visualizations and recommendations for choosing the values of these factors, which make it easier for users to improve approximation accuracy and computational performance. We also propose diagnostics for checking that the number of basis functions and the domain of the prediction space are adequate given the data. The proposed approach is simple and exhibits an attractive computational complexity due to its linear structure, and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\end{abstract}

@@

Gaussian processes (GPs) are flexible statistical models for specifying probability distributions over multi-dimensional non-linear functions \citep{rasmussen2006gaussian,neal1997monte}. Their name stems from the fact that any finite set of function values is jointly distributed as a multivariate Gaussian. GPs are defined by a mean and a covariance function. The covariance function encodes our prior assumptions about the functional relationship, such as continuity, smoothness, periodicity and scale properties. GPs not only allow for non-linear effects but can also implicitly handle interactions between input variables (covariates). Different types of covariance functions can be combined for further increased flexibility. Due to their generality and flexibility, GPs are of broad interest across machine learning and statistics \citep{rasmussen2006gaussian,neal1997monte}. Among others, they find application in the fields of spatial epidemiology \citep{diggle2013statistical,carlin2014hierarchical}, robotics and control \citep{deisenroth2015gaussian}, signal processing \citep{sarkka2013spatiotemporal}, neuroimaging \citep{andersen2017} as well as Bayesian optimization and probabilistic numerics \citep{roberts2010bayesian,briol2015probabilistic,hennig2015probabilistic}.

The key element of a GP is the covariance function that defines the dependence structure between function values at different inputs. However, computing the posterior distribution of a GP comes with a computational issue because of the need of inverting the covariance matrix. Given $n$ observations in the data, the computational complexity and memory requirements of computing the posterior distribution for a GP in general scale as $O(n^3)$ and $O(n^2)$, respectively. This limits their application to rather small data sets of a few tens of thousands observations at most. The problem becomes more severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed.

@@

In this paper, we focus on a recent framework for fast and accurate inference for fully Bayesian GPs using basis function approximations based on approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Using a basis function expansion, a GP is approximated with a linear model which makes inference considerably faster. The linear model structure makes GPs easy to implement as a building block in more complicated models in modular probabilistic programming frameworks, where there is a big benefit if the approximation specific computation is simple. Furthermore, a linear representation of a GP makes it easier to be used as latent function in non-Gaussian observational models allowing for more modelling flexibility. The basis function approximation via Laplace eigenfunctions can be made arbitrary accurate and the trade-off between computational complexity and approximation accuracy can easily be controlled.

@@

We have implemented the approach in the probabilistic programming language Stan \citep{carpenter2017stan} as well as subsequently in the \textit{brms} package \citep{burkner2017brms} of the R software \citep{R2019R}. Several illustrative examples of the performance and applicability of the method are shown using both simulated and real datasets. All examples are accompanied by the corresponding Stan code. Although there are several GP specific software packages available to date, for example, GPML \citep{rasmussen2010gpml},  GPstuff \citep{vanhatalo2013gpstuff}, GPy \citep{gpy2014}, and GPflow \citep{GPflow2017}, each provide efficient implementations only for a restricted range of GP-based models. In this paper, we do not focus on the fastest possible inference for some specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks.

@@

A GP is a stochastic process which defines the distribution over a collection of random variables indexed by a continuous variable, that is, $\left\lbrace f(t): t \in \mathcal{T}\right\rbrace$ for some index set $\mathcal{T}$. GPs have the defining property that the marginal distribution of any finite subset of random variables, $\left\lbrace f(t_1), f(t_2), \hdots, f(t_N) \right\rbrace$, is a multivariate Gaussian distribution.

In this work, GPs will take the role of a prior distribution over function spaces for non-parametric latent functions in a Bayesian setting. Consider a data set $\mathcal{D} = \left\lbrace \bm{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is modelled conditionally as $p(y_n \mid f(\bm{x}_n),\phi)$, where $p$ is some parametric distribution with parameters  $\phi$, and $f$ is an unknown function with GP prior, which depends on an input $\bm{x}_n\in {\rm I\!R}^D$. This generalizes readily to more complex models depending on several unknown functions, for example such as $p(y_n \mid f(\bm{x}_n),g(\bm{x}_n))$ or multilevel models. Our goal is to obtain the posterior distribution for the value of the function $\tilde{f}=f(\tilde{\bm{x}})$  evaluated at a new input point $\tilde{\bm{x}}$.

@@

The approximate GP method, developed by \citet{solin2018hilbert} and further analysed in this paper, is based on considering the covariance operator of a stationary covariance function as a pseudo-differential operator constructed as a series of Laplace operators. Then, the pseudo-differential operator is approximated with \emph{Hilbert space} methods on a compact subset $\Omega \subset {\rm I\!R}^D$ subject to boundary conditions. For brevity, we will refer to these approximate Gaussian processes as HSGPs. Below, we will present the main results around HSGPs relevant for practical applications. More details and mathematical proofs are provided by \citet{solin2018hilbert}. Our starting point for presenting the method is the definition of the covariance function as a series expansion of eigenvalues and eigenfunctions of the Laplacian operator. The mathematical details of this approximation are briefly presented in Appendix~\ref{ch5_app_approx_covfun}.

@@

Within $\Omega$, we can write any stationary covariance function with input values $\{x,x'\} \in \Omega$ as
%
\begin{equation}\label{eq_approxcov}
k(x,x') = \sum_{j=1}^\infty S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x'),
\end{equation} 
%
where $S_{\theta}$ is the spectral density of the stationary covariance function $k$ (see Section \ref{ch4_sec_cov}) and $\theta$ the set of hyperparameters of $k$ \citep{rasmussen2006gaussian}. The terms $\{\lambda_j\}_{j=1}^{\infty}$ and $\{\phi_j(x)\}_{j=1}^{\infty}$ are the sets of eigenvalues and eigenvectors, respectively, of the Laplacian operator in the given domain $\Omega$. Namely, they satisfy the following eigenvalue problem in $\Omega$ when applying the Dirichlet boundary condition (other boundary conditions could be used as well)
%
\begin{align}\label{eq_eigenproblem}
\begin{split}
-\nabla^2 \phi_j(x)&=\lambda \phi_j(x), \hspace{1cm}  x \in \Omega \\ 
\phi_j(x)&= 0, \hspace{1.85cm}   x \notin \Omega.
\end{split}
\end{align} 
%
The eigenvalues $\lambda_j>0$ are real and positive because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalue problem in equation \eqref{eq_eigenproblem} are sinusoidal functions. The solution to the eigenvalue problem is independent of  the specific choice of covariance function and is given by
%
\begin{align}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^{\!2}, \label{eq_eigenvalue}\\
\phi_j(x)&=\sqrt{\frac{1}{L}}\, \sin\!\!\left(\sqrt{\lambda_j}(x+L)\right). \label{eq_eigenfunction}
\end{align}

@@

Thus, the Gram matrix $K$ of the covariance function $k$ for a set of observations $i=1,\ldots,n$ and corresponding input values $\{x_i\}_{i=1}^{n} \in \Omega^{n}$ can be represented as
%
\begin{equation}
 \bm{K} = \Phi \Delta \Phi^\intercal, \nonumber
\end{equation}
%
where $\Phi \in {\rm I\!R}^{n\times m}$ is the matrix of eigenfunctions $\phi_j(x_i)$
%
\begin{align}
\Phi =  \left[ {\begin{array}{ccc}
   \phi_1(x_1) & \cdots & \phi_m(x_1)  \\
    \vdots &\ddots & \vdots  \nonumber \\ 
    \phi_1(x_n) & \cdots & \phi_m(x_n) \\
  \end{array} } \right].
\end{align}
As a result, the model for $f$ can be written as
%
\begin{equation}
\bm{f} \sim \Normal(\bm{\mu},\Phi \Delta \Phi^\intercal). \nonumber
\end{equation}
This equivalently leads to a linear representation of $f$ via
%
\begin{equation}\label{eq_approxf}
f(x) \approx \sum_{j}^m \left( S_{\theta}(\sqrt{\lambda_j})\right)^{\frac{1}{2}} \phi_j(x) \beta_j,
\end{equation}
%
where $\beta_j \sim \Normal(0,1)$. Thus, the function $f$ is approximated with a finite basis function expansion (using the eigenfunctions $\phi_j$ of the Laplace operator), scaled by the square root of spectral density values. A key property of this approximation is that the eigenfunctions $\phi_j$ do not depend on the hyperparameters of the covariance function $\theta$. Instead, the only dependence of the model on $\theta$ is through the spectral density $S_{\theta}$. The eigenvalues $\lambda_j$ are monotonically increasing with $j$ and $S_{\theta}$ goes rapidly to zero for bounded covariance functions. Therefore, eq.~\eqref{eq_approxf} can be expected to be a good approximation for a finite number of $m$ terms in the series as long as the inputs values $x_i$ are not too close to the boundaries $-L$ and $L$ of $\Omega$. The computational cost of evaluating the log posterior density of univariate HSGPs scales as $O(nm + m)$, where $n$ is the number of observations and $m$ the number of basis functions.

>>>>

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions. However, they entail a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis function approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance and practical implementation of the method in relation to key factors such as the number of basis functions, domain of the prediction space, and smoothness of the latent function. We provide intuitive visualizations and recommendations for choosing the values of these factors, which make it easier for users to improve approximation accuracy and computational performance. We also propose diagnostics for checking that the number of basis functions and the domain of the prediction space are adequate given the data. The proposed approach is simple and exhibits an attractive computational complexity due to its linear structure, and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\end{abstract}

@@

Gaussian processes (GPs) are flexible statistical models for specifying probability distributions over multi-dimensional non-linear functions \citep{rasmussen2006gaussian,neal1997monte}. Their name stems from the fact that any finite set of function values is jointly distributed as a multivariate Gaussian distribution. GPs are defined by a mean and a covariance function. The covariance function encodes our prior assumptions about the functional relationship, such as continuity, smoothness, periodicity and scale properties. GPs not only allow for non-linear effects but can also implicitly handle interactions between input variables (covariates). Different types of covariance functions can be combined for further increased flexibility. Due to their generality and flexibility, GPs are of broad interest across machine learning and statistics \citep{rasmussen2006gaussian,neal1997monte}. Among others, they find application in the fields of spatial epidemiology \citep{diggle2013statistical,carlin2014hierarchical}, robotics and control \citep{deisenroth2015gaussian}, signal processing \citep{sarkka2013spatiotemporal}, neuroimaging \citep{andersen2017} as well as Bayesian optimization and probabilistic numerics \citep{roberts2010bayesian,briol2015probabilistic,hennig2015probabilistic}.

The key element of a GP is the covariance function that defines the dependence structure between function values at different inputs. However, computing the posterior distribution of a GP comes with a computational issue because of the need to invert the covariance matrix. Given $n$ observations in the data, the computational complexity and memory requirements of computing the posterior distribution for a GP in general scale as $O(n^3)$ and $O(n^2)$, respectively. This limits their application to rather small data sets of a few tens of thousands observations at most. The problem becomes more severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed.

@@

In this paper, we focus on a recent framework for fast and accurate inference for fully Bayesian GPs using basis function approximations based on approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Using a basis function expansion, a GP is approximated with a linear model which makes inference considerably faster. The linear model structure makes GPs easy to implement as building blocks in more complicated models in modular probabilistic programming frameworks, where there is a big benefit if the approximation specific computation is simple. Furthermore, a linear representation of a GP makes it easier to be used as latent function in non-Gaussian observational models allowing for more modelling flexibility. The basis function approximation via Laplace eigenfunctions can be made arbitrary accurate and the trade-off between computational complexity and approximation accuracy can easily be controlled.

@@

We have implemented the approach in the probabilistic programming language Stan \citep{carpenter2017stan} as well as subsequently in the \textit{brms} package \citep{burkner2017brms} of the R software \citep{R2019R}. Several illustrative examples of the performance and applicability of the method are shown using both simulated and real datasets. All examples are accompanied by the corresponding Stan code. Although there are several GP specific software packages available to date, for example, GPML \citep{rasmussen2010gpml},  GPstuff \citep{vanhatalo2013gpstuff}, GPy \citep{gpy2014}, and GPflow \citep{GPflow2017}, each provide efficient implementations only for a restricted range of GP-based models. In this paper, we do not focus on the fastest possible inference for a small set of specific GP models, but instead we are interested in how GPs can be easily used as modular components in probabilistic programming frameworks.

@@

A GP is a stochastic process which defines the distribution of a collection of random variables indexed by a continuous variable, that is, $\left\lbrace f(t): t \in \mathcal{T}\right\rbrace$ for some index set $\mathcal{T}$. GPs have the defining property that the marginal distribution of any finite subset of random variables, $\left\lbrace f(t_1), f(t_2), \hdots, f(t_N) \right\rbrace$, is a multivariate Gaussian distribution.

In this work, GPs will take the role of a prior distribution over function spaces for non-parametric latent functions in a Bayesian setting. Consider a data set $\mathcal{D} = \left\lbrace (\bm{x}_n, y_n) \right\rbrace_{n=1}^N$, where $y_n$ is modelled conditionally as $p(y_n \mid f(\bm{x}_n),\phi)$, where $p$ is some parametric distribution with parameters  $\phi$, and $f$ is an unknown function with a GP prior, which depends on an input $\bm{x}_n\in {\rm I\!R}^D$. This generalizes readily to more complex models depending on several unknown functions, for example such as $p(y_n \mid f(\bm{x}_n),g(\bm{x}_n))$ or multilevel models. Our goal is to obtain the posterior distribution for the value of the function $\tilde{f}=f(\tilde{\bm{x}})$  evaluated at a new input point $\tilde{\bm{x}}$.

@@

The approximate GP method, developed by \citet{solin2018hilbert} and further analysed in this paper, is based on considering the covariance operator of a stationary covariance function as a pseudo-differential operator constructed as a series of Laplace operators. Then, the pseudo-differential operator is approximated with \emph{Hilbert space} methods on a compact subset $\Omega \subset {\rm I\!R}^D$ subject to boundary conditions. For brevity, we will refer to these approximate Gaussian processes as HSGPs. Below, we will present the main results around HSGPs relevant for practical applications. More details on the theoretical background are provided by \citet{solin2018hilbert}. Our starting point for presenting the method is the definition of the covariance function as a series expansion of eigenvalues and eigenfunctions of the Laplacian operator. The mathematical details of this approximation are briefly presented in Appendix~\ref{ch5_app_approx_covfun}.

@@

Within $\Omega$, we can write any stationary covariance function with input values $x,x' \in \Omega$ as
%
\begin{equation}\label{eq_approxcov}
k(x,x') = \sum_{j=1}^\infty S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x'),
\end{equation} 
%
where $S_{\theta}$ is the spectral density of the stationary covariance function $k$ (see Section \ref{ch4_sec_cov}) and $\theta$ is the set of hyperparameters of $k$ \citep{rasmussen2006gaussian}. The terms $\{\lambda_j\}_{j=1}^{\infty}$ and $\{\phi_j(x)\}_{j=1}^{\infty}$ are the sets of eigenvalues and eigenvectors, respectively, of the Laplacian operator in the given domain $\Omega$. Namely, they satisfy the following eigenvalue problem in $\Omega$ when applying the Dirichlet boundary condition (other boundary conditions could be used as well)
%
\begin{align}\label{eq_eigenproblem}
\begin{split}
-\nabla^2 \phi_j(x)&=\lambda_j \phi_j(x), \hspace{1cm}  x \in \Omega \\ 
\phi_j(x)&= 0, \hspace{1.85cm}   x \notin \Omega.
\end{split}
\end{align} 
%
The eigenvalues $\lambda_j>0$ are real and positive because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalue problem in equation \eqref{eq_eigenproblem} are sinusoidal functions. The solution to the eigenvalue problem is independent of  the specific choice of covariance function and is given by
%
\begin{align}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^{\!2}, \label{eq_eigenvalue}\\
\phi_j(x)&=\sqrt{\frac{1}{L}}\, \sin\!\!\left(\sqrt{\lambda_j}(x+L)\right). \label{eq_eigenfunction}
\end{align}

@@

Thus, the Gram matrix $\bm{K}$ for the covariance function $k$ for a set of observations $i=1,\ldots,n$ and corresponding input values $\{x_i\}_{i=1}^{n}$ can be represented as
%
\begin{equation}
 \bm{K} = \Phi \Delta \Phi^\intercal, \nonumber
\end{equation}
%
where $\Phi \in {\rm I\!R}^{n\times m}$ is the matrix of eigenfunctions $\phi_j(x_i)$
%
\begin{align}
\Phi =  \left[ {\begin{array}{ccc}
   \phi_1(x_1) & \cdots & \phi_m(x_1)  \\
    \vdots &\ddots & \vdots  \nonumber \\ 
    \phi_1(x_n) & \cdots & \phi_m(x_n) \\
  \end{array} } \right].
\end{align}
As a result, the model for $f$ can be written as
%
\begin{equation}
\bm{f} \sim \Normal(\bm{\mu},\Phi \Delta \Phi^\intercal). \nonumber
\end{equation}
This equivalently leads to a linear representation of $f$ via
%
\begin{equation}\label{eq_approxf}
f(x) \approx \sum_{j=1}^m \left( S_{\theta}(\sqrt{\lambda_j})\right)^{\frac{1}{2}} \phi_j(x) \beta_j,
\end{equation}
%
where $\beta_j \sim \Normal(0,1)$. Thus, the function $f$ is approximated with a finite basis function expansion (using the eigenfunctions $\phi_j$ of the Laplace operator), scaled by the square root of spectral density values. A key property of this approximation is that the eigenfunctions $\phi_j$ do not depend on the hyperparameters of the covariance function $\theta$. Instead, the only dependence of the model on $\theta$ is through the spectral density $S_{\theta}$. The eigenvalues $\lambda_j$ are monotonically increasing with $j$ and $S_{\theta}$ goes rapidly to zero for bounded covariance functions. Therefore, eq.~\eqref{eq_approxf} can be expected to be a good approximation for a finite number of $m$ terms in the series as long as the inputs values $x_i$ are not too close to the boundaries $-L$ and $L$ of $\Omega$. The computational cost of evaluating the log posterior density of univariate HSGPs scales as $O(nm + m)$, where $n$ is the number of observations and $m$ the number of basis functions.
